Compilation and Backend-Independent Vectorization for Multi-Party Computation

### 2022-04-29 overleaf/vectorization.tex
Did a pass on Section 5.

\ana{Need an intro here!}

@@

@@

There are three kinds of arrays (for now all kept internally as one-dimensional arrays, but that's under discussion).
\begin{itemize}

\item Scalars: These are scalar variables we lift into arrays for the purposes of vectorization.
For those, all writes are canonical writes and all reads are canonical reads. We will apply raise dimension when
a scalar is used in an inner loop (e.g., {\sf sum!2} in line 6 of the MPC source code will be raised to a 1-dimensional
array since {\sf sum!2} is used in the inner $j$-loop). Drop dimension applies as well; this happens when a scalar
written in an inner loop is used outside of the loop (e.g., {\sf sum!2} for which the lifted inner loop computes
{\sf D} values, but the outer loop only needs the last one.)

\item Read-only input arrays: Read-only inputs. There are NO writes, while we may have non-canonical reads, $f(i,j,k)$.
Phase 1 of Basic vectorization will add raise dimension operation at the beginning of the function to lift these arrays,
and raise dimension may \emph{reshape} arrays. If there are multiple "views" of the input array, there would be multiple raise dimension statements to create
each one of these views. The invariant is that at reads in loops, the reads of "views" of the original input array are canonical.
Only raise dimension applies to these arrays, and only in the beginning of the program. For example, 1-dimensional array
\python{C} is lifted into 2-dimensional array {\sf N x D} by copying the row {\sf N} times.

 \item Read-write output arrays: Writes are canonical (by restriction) but reads can be non-canonical.
 Dependence analysis limits vectorization when non-canonical read access refers to array writes in
 previous iterations, thus creating loop-carried dependences. We may apply both raise and drop dimension, however,
 they respect the fixed dimensionality of the output array. The array cannot be raised to a dimension lower than
 its canonical (fixed) dimensionality and it cannot be dropped lower. In addition, non-canonical reads may
 require lifting (i.e., reshaping) of the array after the most recent write, rather than in the beginning of the program,
 in order to reduce a non-canonical read to a canonical one.

@@

\subsection{Raise Dimension and Drop Dimension}

There are two conceptual versions of $\mathit{raise\_dim}$. One applies on read-only input arrays and reshapes those arrays
when necessary to ensure canonical read access in the corresponding loop. The signature of $\mathit{raise\_dim}$
is as follows. It takes the original array \texttt{C}, the access pattern function $f(i,j,k)$ in loop nest $i,j,k$ and the loop bounds
$((i:I),(j:J),(k:K))$:
\[ \mathit{raise\_dim}(\texttt{A}, f(i,j,k), \texttt{((i:I),(j:J),(k:K))}) \]
It produces a new 3-dimensional array \texttt{A'} by iterating over $i,j,k$ and setting each element of  \texttt{A'} as follows:
\[ \texttt{A'[i,j,k] =  A[f(i,j,k)]} \]
The end result is that uses of \texttt{[A[f(i,jk)]} in loop nest $i,j,k$ are replaced with canonical read-accesses to \texttt{A'[i,j,k]}
that can be vectorized. In the running Biometric example, \texttt{C'} = $\mathit{raise\_dim}$\texttt{(C, j, (i:N,j:D))} lifts the
1-dimensional array \texttt{C} into a 2-dimensional array. The $i,j$ loop now accesses \texttt{C'} in the canonical way, \texttt{C'[i,j]}.
Similarly, \texttt{S'} = $\mathit{raise\_dim}$\texttt{(S, i*D+j, (i:N,j:D))} tries to lift \texttt{S}, but the operation turns into a no-op
because \texttt{S} is already a 2-dimensional array and the read access is canonical.

The other version of $\mathit{raise\_dim}$ applies on scalars and read-write arrays. It lifts a lower-dimension array into a
higher-dimension for access in a nested loop. Here \texttt{A} is an $i$ array and raise dimension adds two additional dimensions:
\[ \mathit{raise\_dim}(\texttt{A}, \texttt{(j:J,k:K)}) \]
This version is reduced to the above version by adding the access pattern function, which is just $i$:
\[ \mathit{raise\_dim}(\texttt{A}, i, \texttt{(j:J,k:K)}) \]

The corresponding $\mathit{drop\_dim}$ is carried out when an array written in an inner loop is used in an enclosing loop.
It takes a higher dimensional array, say $i,j,k$ and removes trailing dimensions, say $j,k$:
\[ \mathit{drop\_dim}(\texttt{A}, \texttt{(j:J,k:K)}) \]
It iterates over $i$ and takes the result at the maximal index of $j$ and $k$, i.e., the result at the last iterations of $j$ and $k$:
\[ \texttt{A'[i] = A[i,J-1,K-1]} \]

@@

@@

\WHILE {there are closure $cl_1$ and $cl_2$ that intersect}
\STATE {merge $cl_1$ and $cl_2$}
\ENDWHILE
\FOR {each closure $cl$ (after merge)}
\STATE {create FOR $i_d = 0; ...$ loop}
\STATE {add $\phi$-nodes in $cl$ to header block}
\STATE {{\color{red}{add target-less $\phi$-node for A if $cl$ updates array A}}}
\STATE {add statements in $cl$ to loop body in some order of dependences}
\STATE \COMMENT { Dimension is not vectorizable: }
\STATE {change ${I_d}$ to $i_d$ in all statements in loop}
\STATE {treat FOR loop as monolith node: some def-use edges become same-level.}
\ENDFOR
{
\color{red}
\FOR {each target-less $\phi$-node $A_1 = \phi(A_0,A_k)$}
\STATE {in vectorizable stmts, replace use of $A_1$ with $A_0$}
\STATE {discard $\phi$-node if not used in any $cl$, replacing $A_1$ with $A_0$ or $A_k$ appropriately}
\ENDFOR
}
\ENDFOR

@@

We build a correctness argument that loosely follows the theory of Abstract Interpretation.
We define the syntax of MPC Source programs. The domain of MPC Source programs
expressible in the syntax (with certain semantic restrictions) is the abstract domain $A$.
We then define the \emph{linearization} of an MPC Source program as an interpretation
over the syntax. The linearization, which is a \emph{schedule}, is the concrete domain $C$.
Since we reason over def-use graphs in $A$ we define a partial order relation over elements of $A$
in terms of def-use relations. We define a partial order over elements of $C$ as well,
in terms of def-use relations in the concrete domain $C$. We prove two theorems
that state (informally) that the schedule corresponding to the original program computes the same
result as the schedule corresponding to the vectorized program.

\begin{figure}%[!b]
%\small
\[
\begin{array}{l@{~~~~~}l}
  \begin{array}{l@{~}l@{~~~}l}
  s &::= s;s & sequence \\
     & \mid \x[i,J,k] = \y[i,J,k] \; \mathit{op\_SIMD} \; \z[i,J,k] & operation \\
     & \mid \x[i,J,k]  = \texttt{PHI}(\x_1[i,J,k],\x_2[i,J,k\!-\!1]) & \mathit{phi} \; \mathit{node} \\
     & \mid \x[i,J,k] = \mathit{raise\_dim}(\x'[i],(J\!:\!J,k\!:\!K)) & \mathit{raise} \; \mathit{dimension(s)} \\
     & \mid \x[i,J] = \mathit{drop\_dim}(\x'[i,J,k],k)  & \mathit{drop} \; \mathit{dimension(s)} \\
     & \mid \x = \y & \mathit{propagation} \\
     & \mid \texttt{FOR} \; 0 \le i < I: s  & \mathit{loop} \\
  \end{array}    \\
\end{array}
\]
\vspace{-3mm}
\caption{MPC Source Syntax}\label{fig:MPC_Source_syntax}
\vspace{-5mm}
\end{figure}

@@

\figref{fig:MPC_Source_syntax} defines the syntax for our intermediate representation, MPC Source.
There are semantics restrictions over the syntax as well: a variable $\x[i,j,k]$ is a 3-dimensional array
$(i:I,j:J,k:K)$ and also, a statement $\x[i,J,k] = ...$ is enclosed in loops over $i$ and $k$ as shown below.
Thus, $i$ and $k$ are in scope.
\begin{verbatim}
FOR 0 <= i < I:
    ...
    FOR 0 <= k < K:
        x[i,J,k] = ...
\end{verbatim}

Statements $\mathit{operation, phi, raise \; dimension(s), drop \; dimension(s)}$ are \emph{base} statements,
and $\mathit{sequence, loop}$ are \emph{compound statements}.

@@

The concretization of the FOR statement is as follows:
\[ \gamma(\texttt{FOR} \; 0 \le i < I: s) = \gamma(s)[0/i] \; ; \; \gamma(s)[1/i] \; ; \; ... \gamma(s)[I-1/i] \]
$\gamma$ simply unrolls the loop substituting $i$ with 0, 1, etc. Here ; denotes sequential execution.

@@

\ana{Write the proof sketch and final argument, etc.}

\subsection{Extension of Basic Vectorization with Array Writes}

@@

{\small
\begin{verbatim}
for i in range(N):
  A[i] = B[i] + 10;
  B[i] = A[i] * D[i-1];
  C[i] = A[i] * D[i-1];
  D[i] = B[i] * C[i];
\end{verbatim}
}

@@

{\small
\begin{verbatim}
for i in range(N):
1. A_1 = PHI(A_0,A_2)
2. B_1 = PHI(B_0,B_2)
3. C_1 = PHI(C_0,C_2)
4. D_1 = PHI(D_0,D_2)
5. A_2 = update(A_1, i, B_1[i] + 10);
6. B_2 = update(B_1, i, A_2[i] * D_1[i-1]);
7. C_2 = update(C_1, i, A_2[i] * D_1[i-1]);
8. D_2 = update(D_1, i, B_2[i] * C_2[i]);
\end{verbatim}
}

There is a cycle around \texttt{B\_1 = PHI(B\_0,B\_2)} that includes statement
\texttt{A\_1 = update(A\_0, i, B\_1[i] + 10)} and that statement won't be vectorized even
though in fact there is no loop-carried dependency from the write of \texttt{B\_1[i]} at 6 to the
read of \texttt{ ... = B\_1[i]} at 8.

@@

\begin{algorithmic}
\FOR {each array $\texttt{A}$ written in loop $j$}
\STATE \COMMENT { including enclosed loops in $j$ }
\STATE dep = False
\FOR {each pair def: $\texttt{A}_m[f(i,j,k)] = ... $, and use: $ ... = \texttt{A}_n[f'(i,j,k)]$ in loop $j$}
\IF {$\exists \underline{i}, \underline{j}, \underline{j}', \underline{k}, \underline{k}'$, s.t. $0\! \le\! \underline{i}\! <\! I,\: 0\! \le\! \underline{j}, \underline{j}' \! <\! J,\: 0\! \le\! \underline{k}, \underline{k}' \! <\! K,\: \underline{j}<\underline{j}'$, and $f(\underline{i},\underline{j},\underline{k}) = f'(\underline{i},\underline{j}',\underline{k}')$}
\STATE dep = True
\ENDIF
\ENDFOR
\IF {dep == False}
\STATE remove back edge into $\texttt{A}$'s $\phi$-node in loop $j$.
\ENDIF
\ENDFOR
\end{algorithmic}

Consider a loop $j$ enclosed in some fixed $\underline{i}$. Only if an update (definition) $\texttt{A}_m[f(i,j,k)] = ... $ at some iteration $\underline{j}$
references the \emph{same} array element as a use $ ... = \texttt{A}_n[f'(i,j,k)]$ at some later iteration $\underline{j}'$,
we may have a loop-carried dependence for $\texttt{A}$ due to this def-use pair. (In contrast, Cytron's algorithm inserts a loop-carried dependency every time there is an array update.)
The algorithm above examines all def-use pairs in loop $j$, including defs and uses in nested loops, searching for values $\underline{i}, \underline{j}, \underline{j}', \underline{k}, \underline{k}'$ that satisfy
$f(\underline{i},\underline{j},\underline{k}) = f'(\underline{i},\underline{j}',\underline{k}')$. If such values exist for some def-use pair, then there is a potential
loop-carried dependence on $\texttt{A}$; otherwise there is not and we can remove the spurious backward edge thus ``freeing up'' statements for vectorization. %\ana{Double check this!!!}

Consider the earlier example. There is a single loop, $i$. Clearly, there is no pair $\underline{i}$ and $\underline{i}'$, where $\underline{i} < \underline{i}'$ that make $\underline{i} = \underline{i}'$
due to the def-use pairs of \texttt{A} 5-6 and 5-7.
Therefore, we remove the back edge from 5 to the phi-node 1. Analogously, we remove the back edges from 6 to 2 and from 7 to 3. However, there are many values $\underline{i} < \underline{i'}$ that make $\underline{i} = \underline{i'}-1$ and the back edge from 8 to 4 remains (def-use pairs for \texttt{D}). As a result of removing these spurious edges, Basic Vectorization will find that statement 5 is vectorizable. Statements 6, 7 and 8 will correctly appear in the FOR loop.

Note however, that this step renders some array phi-nodes target-less. We handle target-less phi-nodes with a minor extension of Basic Vectorization (Phase 2).
First, we merge closures that update the same array. This simplifies handling of array $\phi$-nodes: if each closure is turned into a separate loop
each loop will need to have its own array phi-node to account for the update and this would complicate the analysis.
Second, we add the target-less node of array \texttt{A} back to the closure that updates \texttt{A} ---
the intuition is, even if there is no loop-carried dependence from writes to reads on \texttt{A}, \texttt{A} is written and the write (i.e., update) cannot be vectorized;
therefore, the updated array has to carry to the next iteration of the loop. Third, in cases when the phi-node remains target-less, i.e., cases when the
array write can be vectorized, we have to properly remove the phi-node replacing uses of the left-hand side of the phi-node with its arguments.

@@

For now, we restrict array updates to \emph{canonical updates}. Assume (for simplicity) a two-dimensional array \texttt{A[I,J]}. A canonical update is the following:
\begin{verbatim}
for i in range(I):
  for j in range(J):
    ...
    A[i,j] = ...
    ...
\end{verbatim}

>>>>

\ana{Need an intro/overview here!}
\ana{Need to define notion of canonical access!}
\ana{Talk about MPC Source, opportunities for vectorization, algorithm unique to MPC Source.}

@@

\subsubsection{Overview}

@@

There are three kinds of arrays.
\begin{itemize}

\item Scalars: These are scalar variables we expand into arrays for the purposes of vectorization.
For those, all writes are canonical writes and all reads are canonical reads. We will \emph{raise dimension} when
a scalar gives rise to an outer-to-inner dependence edge (e.g., {\sf sum!2} in line 6 of the MPC source code will be raised to a 1-dimensional
array since {\sf sum!2} is used in the inner $j$-loop). We will \emph{drop dimension} when a scalar
gives rise to an inner-to-outer dependence edge (e.g., {\sf sum!2} for which the lifted inner loop computes
{\sf D} values, but the outer loop only needs the last one.)

\item Read-only input arrays: There are no writes, while we may have non-canonical reads, $f(i,j,k)$.
Vectorization adds raise dimension operations at the beginning of the function to lift these arrays to 
the dimensionality of the loop where they are used, possibly \emph{reshaping} the arrays. 
If there are multiple ``views'' of the input array, there would be multiple raise dimension statements to create
each one of these views. The invariant is that at reads in loops, the reads of "views" of the original input array are canonical.
%Only raise dimension applies to these arrays, and only in the beginning of the program. For example, 1-dimensional array
%\python{C} is lifted into 2-dimensional array {\sf N x D} by copying the row {\sf N} times.

 \item Read-write output arrays: Writes are canonical (by restriction) but reads can be non-canonical.
 Dependences limit vectorization when non-canonical read access refers to array writes in
 previous iterations, thus creating loop-carried dependences. We may apply both raise and drop dimension, however,
 they respect the fixed dimensionality of the output array. The array cannot be raised to a dimension lower than
 its canonical (fixed) dimensionality and it cannot be dropped to lower dimension. In addition, non-canonical reads may
 require lifting (i.e., reshaping) of the array after the most recent write rather than in the beginning of the program
 in order to reduce a non-canonical read to a canonical one.

@@

\subsubsection{Raise Dimension and Drop Dimension}

There are two conceptual versions of $\mathit{raise\_dim}$. One reshapes an arbitrary access into a canonical read access in the corresponding loop. 
%The signature of $\mathit{raise\_dim}$ is as follows. 
It takes the original array, the access pattern function $f(i,j,k)$ in loop nest $i,j,k$ and the loop bounds
$((i:I),(j:J),(k:K))$:
\[ \mathit{raise\_dim}({\sf A, f(i,j,k), ((i:I),(j:J),(k:K))}) \]
It produces a new 3-dimensional array {\sf A'} by iterating over $i,j,k$ and setting each element of  {\sf A'} as follows:
\[ {\sf A'[i,j,k] =  A[f(i,j,k)]} \]
The end result is that uses of {\sf [A[f(i,j,k)]} in loop nest $i,j,k$ are replaced with canonical read-accesses to {\sf A'[i,j,k]}
that can be vectorized. In the running Biometric example, {\sf C'} = $\mathit{raise\_dim}${\sf (C, j, (i:N,j:D))} lifts the
1-dimensional array {\sf C} into a 2-dimensional array. The $i,j$ loop now accesses {\sf C'} in the canonical way, {\sf C'[i,j]}.
Similarly, {\sf S'} = $\mathit{raise\_dim}${\sf (S, i*D+j, (i:N,j:D))} tries to lift {\sf S}, but the operation turns into a no-op
because {\sf S} is already a 2-dimensional array and the read access is canonical.

The other version of $\mathit{raise\_dim}$ lifts a lower-dimension array into a
higher-dimension for access in a nested loop. It is necessary when processing outer-to-inner dependences. 
Here {\sf A} is an $i$-array and raise dimension adds two additional dimensions:
\[ \mathit{raise\_dim}({\sf A}, {\sf (j:J,k:K)}) \]
This version is reduced to the above version by adding the access pattern function, which is just $i$:
\[ \mathit{raise\_dim}({\sf A, i, (j:J,k:K)}) \]

The corresponding $\mathit{drop\_dim}$ is carried out when an array written in an inner loop is used in an enclosing loop.
It takes a higher dimensional array, say $i,j,k$ and removes trailing dimensions, say $j,k$:
\[ \mathit{drop\_dim}({\sf A, (j:J,k:K)}) \]
It iterates over $i$ and takes the result at the maximal index of $j$ and $k$, i.e., the result at the last iterations of $j$ and $k$:
\[ {\sf A'[i] = A[i,J-1,K-1]} \]

@@

We present our vectorization algorithm in detail followed by a detailed example on Biometric Matching.

\subsubsection{Algorithm}

There are two key phases of the algorithm. Phase 1 applies dependence analysis and inserts
raise dimension and drop dimension operations as necessary. After this phase operations work
on arrays of the corresponding dimensionality and we optimistically vectorize all arrays.

Phase 2 takes into account dependence analysis. It proceeds from the inner-most towards the outer-most 
loop. For each loop it anchors dependence cycles (closures) around pseudo PHI nodes, removing vectorization 
from the dimension of the loop.

There are several important points in this phase. We may break a loop into smaller loops which would discover opportunities 
for optimization in intermediate statements in the loop. We remove redundant back edges into PHI nodes that are 
due entirely to arrays, which opens up opportunities as well. The pseudo-code in red deals with this extension
of the basic algorithm. We determine that some array pseudo-PHIs are \emph{target-less}, meaning
that there is no loop-carried dependence for that array and back-edges into these nodes can be safely removed. 
This opens up opportunities for vectorization across the write array. Handling of target-less nodes is non-trivial, 
as in certain cases the node cannot be removed. \secref{sec:extension_with_writes} details our algorithm.

\ana{Add an example from histogram here.}

Phases 3 and 4 are less important --- phase 3 cleans up local arrays of references and phase 4 turns operations into MOTION operations.

@@

\WHILE {there are closure $cl_1$ and $cl_2$ that intersect}
\STATE {merge $cl_1$ and $cl_2$}
\ENDWHILE
\FOR {each closure $cl$ (after merge)}
\STATE {create FOR $i_d = 0; ...$ loop}
\STATE {add $\phi$-nodes in $cl$ to header block}
\STATE {{\color{red}{add target-less $\phi$-node for A if $cl$ updates array A}}}
\STATE {add statements in $cl$ to loop body in some order of dependences}
\STATE \COMMENT { Dimension is not vectorizable: }
\STATE {change ${I_d}$ to $i_d$ in all statements in loop}
\STATE {treat FOR loop as monolith node for def-uses: e.g., some def-use edges become same-level.}
\ENDFOR
{
\color{red}
\FOR {each target-less $\phi$-node $A_1 = \phi(A_0,A_k)$}
\STATE {in vectorizable stmts, replace use of $A_1$ with $A_0$}
\STATE {discard $\phi$-node if not used in any $cl$, replacing $A_1$ with $A_0$ or $A_k$ appropriately}
\ENDFOR
}
\ENDFOR

@@

We build a correctness argument that loosely follows Abstract Interpretation.
First we define the MPC Source syntax. The domain of MPC Source programs
expressible in the syntax (with certain semantic restrictions) is the abstract domain $A$.
We then define the \emph{linearization} of an MPC Source program as an interpretation
over the syntax. The linearization, which is a \emph{schedule} (as in~\secref{sec:overview}), is the concrete domain $C$.
Since we reason over def-use graphs in $A$ we define a partial order relation over elements of $A$
in terms of def-use relations. We define a partial order over elements of $C$ as well,
in terms of def-use relations in the concrete domain $C$. We prove two theorems
that state (informally) that the schedule corresponding to the original program computes the same
result as the schedule corresponding to the vectorized program.

\begin{comment}
\begin{figure}%[!b]
%\small
\[
\begin{array}{l@{~~~~~}l}
  \begin{array}{l@{~}l@{~~~}l}
  s &::= s;s & sequence \\
     & \mid \x[i,J,k] = \y[i,J,k] \; \mathit{op\_SIMD} \; \z[i,J,k] & operation \\
     & \mid \x[i,J,k]  = \texttt{PHI}(\x_1[i,J,k],\x_2[i,J,k\!-\!1]) & \mathit{phi} \; \mathit{node} \\
     & \mid \x[i,J,k] = \mathit{raise\_dim}(\x'[i],(J\!:\!J,k\!:\!K)) & \mathit{raise} \; \mathit{dimension(s)} \\
     & \mid \x[i,J] = \mathit{drop\_dim}(\x'[i,J,k],k)  & \mathit{drop} \; \mathit{dimension(s)} \\
%     & \mid \x = \y & \mathit{propagation} \\
     & \mid \texttt{FOR} \; 0 \le i < I: s  & \mathit{loop} \\
  \end{array}    \\
\end{array}
\]
\vspace{-3mm}
\caption{MPC Source Syntax. \ana{Add more to caption.}}\label{fig:MPC_Source_syntax}
\vspace{-5mm}
\end{figure}
\end{comment}

\begin{figure*}%[!b]
\[
\begin{array}{l@{~~~~~}l}
  \begin{array}{l@{~~~~~~~~~}l@{~~~~~~~~~}l@{~~~~~~~~~}l}
  s &::= s_1;s_2 & \gamma(s) = \gamma(s_1) \; ; \; \gamma(s_2) & \mathit{sequence} \\
     & \mid \x[i,J,k] = {\sf op\_SIMD}(\y[i,J,k], \z[i,J,k]) & \gamma(\x[i,J,k] = {\sf op\_SIMD}(\y[i,J,k], \z[i,J,k])) = & \mathit{operation} \\ 
     &                                                                               & ~~\x[i,0,k] = \y[i,0,k] \; {\sf op} \; \z[i,0,k] \;\; || &  \\
     &                                                                               & ~~~~\x[i,1,k] = \y[i,1,k] \; {\sf op} \; \z[i,1,k] \;\; || \; ...\; || & \\
     &                                                                               & ~~~~~~\x[i,J\!-\!1,k] = \y[i,J\!-\!1,k] \; {\sf op} \; \z[i,J\!-\!1,k] & \\
     & \mid \x[i,J,k] = {\sf const} & \mbox{analogous} & \mathit{constant} \\
     & \mid \x[i,J,k]  = {\sf PHI}(\x_1[i,J,k],\x_2[i,J,k\!-\!1]) & & \mathit{pseudo} \; \mathit{PHI} \\
     & \mid \x[i,J,k] = {\sf raise\_dim}(\x'[i],(J\!:\!J,k\!:\!K)) & & \mathit{raise} \; \mathit{dimension(s)} \\
     & \mid \x[i,J] = {\sf drop\_dim}(\x'[i,J,k],k) &  & \mathit{drop} \; \mathit{dimension(s)} \\
%     & \mid \x = \y & & \mathit{propagation} \\
     & \mid {\sf for} \; i \; {\sf in} \; {\sf range}(I): s  & \gamma({\sf for} \; i \; {\sf in} \; {\sf range}(I): s) = & \mathit{loop} \\
     & & ~~\gamma(s)[0/i] \; ; \; \gamma(s)[1/i] \; ; \; ... \; ; \; \gamma(s)[I\!-\!1/i] & \\
  \end{array}    \\
\end{array}
\]
\vspace{-3mm}
\caption{MPC Source Syntax and Semantics. $\gamma$ defines the semantics of MPC source which is a linearization of MPC Source. A SIMD operation parallelizes operations across the vectorized $J$ dimension. $||$ denotes parallel execution, which is standard. $\gamma$ of a for loop unrolls the loop. $;$ denotes sequential execution. Iterative MPC Source trivially extends to non-vectorized dimensions over the enclosing loops.}\label{fig:MPC_Source_syntax}
\label{fig:mpc_syntax}
\vspace{-5mm}
\end{figure*}

@@

\figref{fig:mpc_syntax} states the syntax and linearization semantics of MPC Source. For simiplicity,
we consider only scalars and read-only arrays, however, the treatment extends to write arrays as well.
$\x[i,J,k]$ denotes the value of scalar variable \x{} at loop nest $i,j,k$. 
%(We use 3 dimensions, $i,j,k$, however, the treatment generalizes to arbitrary dimensions.) 
Upper case $J$ denotes a vectorized dimension and lower case $i,k$ denote iterative dimensions. Our compiler imposes 
semantic restrictions over the syntax: (1) $\x$ is treated as a 3-dimensional array and (2) $\x[i,J,k]$ must be enclosed
into for-loops on non-vectorized dimensions $i$ and $k$:
\begin{pythonn}
for i in range(I):
   ...
   for k in range(K):
      ... x[i,J,k] ...
\end{pythonn}

\begin{comment}
\figref{fig:MPC_Source_syntax} defines the syntax for our intermediate representation, MPC Source.
Note that this is a generalization of MPC Source to include vectorized MPC Source.
There are semantic restrictions over the syntax as well: a variable $\x[i,j,k]$ is a 3-dimensional array
$(i:I,j:J,k:K)$ and also, a statement $\x[i,J,k] = ...$ is enclosed in loops over the non-vectorized dimensions 
$i$ and $k$ as shown below. Thus, $i$ and $k$ are in scope.
\begin{pythonn}
for i in range(I):
    ...
    for k in range(K):
        x[i,J,k] = ...
\end{pythonn}

%Statements $\mathit{operation, phi, raise \; dimension(s), drop \; dimension(s)}$ are \emph{base} statements,
%and $\mathit{sequence, loop}$ are \emph{compound statements}.

@@

The concretization of the FOR statement is as follows:
\[ \gamma(\texttt{FOR} \; 0 \le i < I: s) = \gamma(s)[0/i] \; ; \; \gamma(s)[1/i] \; ; \; ... \gamma(s)[I-1/i] \]
$\gamma$ simply unrolls the loop substituting $i$ with 0, 1, etc. Here ; denotes sequential execution.
\end{comment}

\begin{comment}
\paragraph{Partial Orders}

@@

\ana{Write the proof sketch and final argument, etc.}
\end{comment}

For each MPC Source program $a$ we compute the def-use edges in the standard way: if base statement $s1 \in a$ defines variable \x, e.g., $\x[i,j,k] = ...$, 
and base statement $s2 \in a$ uses \x, e.g., $... = ... \x[i,j,k]$ and there is a path in CFG from $s1$ to $s2$, then there is a def-use edge from
 $s1$ to $s2$. We extend the dimensionality of a statement into $s1[i,j,k]$ where $s1[i,j,k]$ inherits the dimensionality of the left-hand-side of the assignment.

 \ana{Add on the linearization semantics?}

 Let $a_0, a_1$ be two MPC Source programs in $A$. 
 Two statements, $s \in a_0$ and $s' \in a_1$ are \emph{same}, written $s \equiv s'$ if they are of the same operation and they operate on 
 the same variables: same variable name and same dimensionality. Recall that dimensions in MPC Source are either iterative (lower case), or vectorized (upper case). 
 Two statements are same even if one operates on an iterative dimension and the other one operates on a vectorized one, e.g., $s[i,j,k] \equiv s'[I,j,K]$.
 We extend the definition to def-use edges in the obvious way: a def-use edge $s_0 \rightarrow s_1$ in $a_0$ and an edge $s'_0 \rightarrow s'_1$ in $a_1$
 are \emph{same}, written $s_0 \rightarrow s_1 \equiv s'_0 \rightarrow s'_1$, if and only if $s_0 \equiv s'_0$, $s_1 \equiv s'_1$, and the two edges are both either 
 forward or backward.

\begin{definition} Let $a_0, a_1 \in A$. We say that $a_0 \le a_1$ iff for every def-use edge $e$ in $a_0$ there is an edge
$e'$ in $a_1$ such that $e \equiv e'$.
\end{definition}

 The def-use edges in the concrete schedule are as expected: there is a def-use edge from statement $s1$ that defines $\x[\underline{i},\underline{j},\underline{k}]$
 to statement $s2$ that uses $\x[\underline{i},\underline{j},\underline{k}]$ if $s1$ is scheduled ahead of $s2$ in the linear schedule. We note that the underlined
 indices, e.g., $\underline{i}$, refer to fixed values, not iterative or vectorized dimensions since in the concrete schedule all induction variables are expanded. 
 E.g., there is a def-use edge from the statement that defines $\x[0,1,2]$ and a statement that uses $\x[0,1,2]$.

\paragraph{Theorems} The two theorems arising from the Basic vectorization optimization are as follows:

\begin{theorem} Let $a_0$ be the iterative MPC Source and let $a_1$ be the vectorized MPC Source computed by Basic vectorization. 
We have that $a_0 \le a_1$. 
\end{theorem}

Theorem 1 states that a def-use edge in MPC Source $a_0$ gives rise to a set of def-use edges in $\gamma(a_0)$
and a ``same'' edge in $a_1$ gives rise to a set of def-use edges in $\gamma(a_1)$ and the former set is included in the latter one.
This means that the computation of $\gamma(a_0)$ is carried out by $\gamma(a_1)$ as well, computing the same result. 
The theorem allows that $a_1$ and $\gamma(a_1)$ have larger state that $a_0$ and $\gamma(a_0)$, i.e., doing additional 
computation into additional memory locations. Theorem 2 establishes the correctness of vectorization --- 
clearly, vectorization preserves all statements and def-use edges in the original MPC Source, and therefore the linearization computes the 
same result as the original MPC Source. (In this case we have an equivalence, $a_0 = a_1$ according to our partial order.)

\subsection{Extension with Array Writes}
\label{sec:extension_with_writes}

@@

{\small
\begin{pythonn}
for i in range(N):
  A[i] = B[i] + 10;
  B[i] = A[i] * D[i-1];
  C[i] = A[i] * D[i-1];
  D[i] = B[i] * C[i];
\end{pythonn}
}

@@

{\small
\begin{pythonn}
for i in range(N):
  A_1 = PHI(A_0,A_2)
  B_1 = PHI(B_0,B_2)
  C_1 = PHI(C_0,C_2)
  D_1 = PHI(D_0,D_2)
  A_2 = update(A_1, i, B_1[i] + 10);
  B_2 = update(B_1, i, A_2[i] * D_1[i-1]);
  C_2 = update(C_1, i, A_2[i] * D_1[i-1]);
  D_2 = update(D_1, i, B_2[i] * C_2[i]);
\end{pythonn}
}

There is a cycle around {\sf B\_1 = PHI(B\_0,B\_2)} that includes statement
{\sf A\_1 = update(A\_0, i, B\_1[i] + 10)} and that statement won't be vectorized even
though in fact there is no loop-carried dependency from the write of {\sf B\_1[i]} at 7 to the
read of {\sf ... = B\_1[i]} at 6.

@@

\begin{algorithmic}
\FOR {each array ${\sf A}$ written in loop $j$}
\STATE \COMMENT { including enclosed loops in $j$ }
\STATE dep = False
\FOR {each pair def: ${\sf A}_m[f(i,j,k)] = ... $, and use: $ ... = {\sf A}_n[f'(i,j,k)]$ in loop $j$}
\IF {$\exists \underline{i}, \underline{j}, \underline{j}', \underline{k}, \underline{k}'$, s.t. $0\! \le\! \underline{i}\! <\! I,\: 0\! \le\! \underline{j}, \underline{j}' \! <\! J,\: 0\! \le\! \underline{k}, \underline{k}' \! <\! K,\: \underline{j}<\underline{j}'$, and $f(\underline{i},\underline{j},\underline{k}) = f'(\underline{i},\underline{j}',\underline{k}')$}
\STATE dep = True
\ENDIF
\ENDFOR
\IF {dep == False}
\STATE remove back edge into ${\sf A}$'s $\phi$-node in loop $j$.
\ENDIF
\ENDFOR
\end{algorithmic}

Consider a loop $j$ enclosed in some fixed $\underline{i}$. Only if an update (definition) ${\sf A}_m[f(i,j,k)] = ... $ at some iteration $\underline{j}$
references the \emph{same} array element as a use $ ... = {\sf A}_n[f'(i,j,k)]$ at some later iteration $\underline{j}'$,
we may have a loop-carried dependence for ${\sf A}$ due to this def-use pair. (In contrast, Cytron's algorithm inserts a loop-carried dependency every time there is an array update.)
The algorithm above examines all def-use pairs in loop $j$, including defs and uses in nested loops, searching for values $\underline{i}, \underline{j}, \underline{j}', \underline{k}, \underline{k}'$ that satisfy
$f(\underline{i},\underline{j},\underline{k}) = f'(\underline{i},\underline{j}',\underline{k}')$. If such values exist for some def-use pair, then there is a potential
loop-carried dependence on ${\sf A}$; otherwise there is not and we can remove the spurious backward edge thus ``freeing up'' statements for vectorization. %\ana{Double check this!!!}

Consider the earlier example. There is a single loop, $i$. Clearly, there is no pair $\underline{i}$ and $\underline{i}'$, where $\underline{i} < \underline{i}'$ that make $\underline{i} = \underline{i}'$
due to the def-use pairs of {\sf A} 6-7 and 6-8.
Therefore, we remove the back edge from 6 to the phi-node 2. Analogously, we remove the back edges from 7 to 3 and from 8 to 4. However, there are many values $\underline{i} < \underline{i'}$ that make $\underline{i} = \underline{i'}-1$ and the back edge from 9 to 5 remains (def-use pairs for {\sf D}). As a result of removing these spurious edges, Vectorization will find that statement 6 is vectorizable. Statements 7, 8 and 9 will correctly appear in the FOR loop.

Note however, that this step renders some array phi-nodes target-less. We handle target-less phi-nodes with a minor extension of Vectorization (Phase 2).
First, we merge closures that update the same array. This simplifies handling of array $\phi$-nodes: if each closure is turned into a separate loop
each loop will need to have its own array phi-node to account for the update and this would complicate the analysis.
Second, we add the target-less node of array {\sf A} back to the closure that updates {\sf A} ---
the intuition is, even if there is no loop-carried dependence from writes to reads on {\sf A}, {\sf A} is written and the write (i.e., update) cannot be vectorized;
therefore, the updated array has to carry to the next iteration of the loop. Third, in cases when the phi-node remains target-less, i.e., cases when the
array write can be vectorized, we have to properly remove the phi-node replacing uses of the left-hand side of the phi-node with its arguments.

@@

For now, we restrict array updates to \emph{canonical updates}. Assume (for simplicity) a two-dimensional array {\sf A[I,J]}. A canonical update is the following:
\begin{pythonn}
for i in range(I):
  for j in range(J):
    ...
    A[i,j] = ...
    ...
\end{pythonn}
